\documentclass[10pt,twocolumn,a4paper]{article}

\usepackage[a4paper,top=2.5cm,bottom=2.5cm,left=2.0cm,right=2.0cm,columnsep=0.65cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{xcolor}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{caption}
\usepackage{float}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{titlesec}

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,
                decorations.pathreplacing,calc,patterns}

\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalsize\bfseries\itshape}{\thesubsubsection}{1em}{}
\titlespacing*{\section}{0pt}{10pt}{5pt}
\titlespacing*{\subsection}{0pt}{7pt}{3pt}

\captionsetup{font=small,labelfont=bf,labelsep=period}

\tikzset{
  block/.style={rectangle,draw=blue!60,fill=blue!10,very thick,minimum height=1.0em,
    minimum width=3em,text centered,rounded corners=2pt,font=\scriptsize},
  blockG/.style={rectangle,draw=green!60!black,fill=green!10,very thick,minimum height=1.0em,
    minimum width=3em,text centered,rounded corners=2pt,font=\scriptsize},
  blockR/.style={rectangle,draw=red!60,fill=red!10,very thick,minimum height=1.0em,
    minimum width=3em,text centered,rounded corners=2pt,font=\scriptsize},
  blockO/.style={rectangle,draw=orange!70,fill=orange!10,very thick,minimum height=1.0em,
    minimum width=3em,text centered,rounded corners=2pt,font=\scriptsize},
  blockP/.style={rectangle,draw=purple!60,fill=purple!10,very thick,minimum height=1.0em,
    minimum width=3em,text centered,rounded corners=2pt,font=\scriptsize},
  arrow/.style={-{Stealth[length=5pt]},very thick,draw=gray!70},
  darrow/.style={<->,very thick,draw=gray!70},
}

\begin{document}

\twocolumn[{%
\begin{center}
{\LARGE\bfseries From X-ray to CT: A Comprehensive Survey of 2D-to-3D Reconstruction Methods From Classical Statistical Shape Models to Generative and Neural Implicit Frameworks\par}
\vspace{0.9em}

{\large Sheetal Lodhi\par}
\vspace{0.4em}

{\small
Department of Computer Science \& Engineering, Netaji Subhas University of Technology, New Delhi, India\par
\texttt{sheetallodhi28@gmail.com}\par}
\vspace{1.0em}
\end{center}

\noindent\textbf{Abstract}\par
\vspace{0.3em}
\noindent
The reconstruction of three-dimensional (3D) volumetric anatomical structures from two-dimensional (2D) X-ray images is a fundamental challenge in medical imaging, with profound implications for clinical diagnosis, surgical planning, and patient-specific treatment. While Computed Tomography (CT) remains the gold standard for 3D anatomical visualization, its high radiation dose, cost, and limited accessibility have driven decades of research toward deriving volumetric information from low-dose, widely available X-ray projections. This survey provides a comprehensive and chronologically organized review of 2D-to-3D reconstruction methods spanning from classical statistical shape models and geometric registration techniques through early convolutional neural networks (CNNs), to the modern era of generative adversarial networks (GANs), diffusion models, and neural radiance fields (NeRF). Building upon the foundational taxonomy established by Maken and Gupta (2023), which systematically reviewed methods up to approximately 2020, this paper extends the timeline by critically analyzing the subsequent wave of data-driven deep generative frameworks. We examine landmark GAN-based methods including CVAE-GAN, X2CT-GAN, TRCT-GAN, DP-GAN+B, AP2CT-GAN, and PTX2CT-GAN, highlighting architectural innovations and clinical targets. We document the transition toward score-based diffusion models such as DX2CT and DVG-Diffusion, and review neural radiance field approaches including MedNeRF and X2BR. For each category, we analyze architectural design choices, datasets, evaluation metrics, anatomical scope, and clinical implications. Finally, we identify persistent challenges and outline promising research directions.

\vspace{0.5em}
\noindent\textbf{Keywords:} X-ray to CT reconstruction; 2D-to-3D reconstruction; generative adversarial networks; diffusion models; neural radiance fields; medical image synthesis; deep learning; statistical shape models; volumetric reconstruction

\vspace{0.8em}
\hrule
\vspace{1.0em}
}]

%% =========================================================
\section{Introduction}
\label{sec:intro}
%% =========================================================

The human body is an inherently three-dimensional anatomical structure.
Accurate visualization of internal anatomy is indispensable for modern clinical
practice, encompassing disease diagnosis, preoperative planning, intraoperative
guidance, and post-treatment follow-up~\cite{maken2023}. Computed Tomography
(CT) acquires hundreds of X-ray projections from different angles and
reconstructs a volumetric image using filtered back-projection (FBP) or
iterative reconstruction. The resulting volumetric data provides unambiguous
information about internal structures, tissue densities, and spatial
relationships.

However, CT imaging carries significant drawbacks. The radiation dose associated
with a single CT examination can be 100 to 500 times that of a conventional
X-ray, raising substantial concerns about cumulative radiation exposure,
particularly for pediatric patients, pregnant women, and those requiring
repeated imaging. CT scanners are also expensive, bulky, and unavailable in
many resource-limited healthcare settings~\cite{gupta2020}.

Plain X-ray radiography is the most widely deployed medical imaging modality
worldwide. X-ray machines are inexpensive, portable, deliver orders-of-magnitude
lower radiation doses, and produce images within seconds. Their fundamental
limitation, however, is that an X-ray image is a two-dimensional projection of a
three-dimensional structure. Overlapping anatomy, loss of depth information, and
magnification distortion all limit its diagnostic utility in complex clinical
scenarios~\cite{maken2023}.

This central tension---between the safety and accessibility of X-rays and the
diagnostic richness of CT---has motivated decades of research into methods for
reconstructing 3D volumetric information from 2D X-ray images. The field has
evolved through four distinct technological eras, as illustrated in
Fig.~\ref{fig:timeline}:
\begin{enumerate}[leftmargin=*,topsep=2pt,itemsep=1pt]
\item \textbf{Classical era (1990--2010):} Statistical shape models (SSMs),
  parametric anatomical models, point-based stereo correspondence, contour-based
  template deformation, and intensity-based registration~\cite{maken2023}.
\item \textbf{Early CNN era (2016--2020):} Convolutional neural networks that
  learn complex 2D-to-3D mappings directly from paired training
  data~\cite{kasten2020,dixit2019,kim2019}.
\item \textbf{GAN era (2019--2023):} Generative adversarial networks producing
  sharper, more realistic volumetric outputs~\cite{x2ctgan,trctgan,dpganb}.
\item \textbf{Diffusion / NeRF era (2022--present):} Score-based diffusion models
  and neural radiance fields enabling high-fidelity sparse-view
  reconstruction~\cite{dx2ct,dvgdiffusion,mednerf,x2br}.
\end{enumerate}

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.80,every node/.style={font=\scriptsize}]
  \draw[very thick,-{Stealth[length=7pt]}] (0,0) -- (8.6,0)
    node[right,font=\small] {\textbf{Year}};
  \fill[blue!12] (0,-0.4) rectangle (3.2,0.4);
  \fill[green!12] (3.2,-0.4) rectangle (5.0,0.4);
  \fill[orange!12] (5.0,-0.4) rectangle (6.8,0.4);
  \fill[red!12] (6.8,-0.4) rectangle (8.5,0.4);
  \foreach \x/\yr in {0/1990,1.6/2000,3.2/2010,4.1/2016,5.0/2019,5.9/2021,6.8/2023,7.7/2025}{
    \draw (\x,-0.45) -- (\x,0.45);
    \node[below] at (\x,-0.45) {\yr};
  }
  \node[blue!70!black,font=\scriptsize\bfseries] at (1.6,0)
    {Classical / SSM};
  \node[green!60!black,font=\scriptsize\bfseries] at (4.1,0)
    {CNN};
  \node[orange!70!black,font=\scriptsize\bfseries] at (5.9,0)
    {GAN};
  \node[red!60!black,font=\scriptsize\bfseries] at (7.65,0)
    {Diff./NeRF};
\end{tikzpicture}
\caption{Chronological evolution of 2D X-ray to 3D CT reconstruction methods
across four distinct eras.}
\label{fig:timeline}
\end{figure}

The specific contributions of this paper are: (1) a unified recapitulation of
classical and early-CNN methods establishing a common chronological baseline;
(2) a structured review of GAN-based methods with detailed architecture
diagrams; (3) analysis of the transition from GANs to diffusion models;
(4) a review of neural radiance field approaches; and (5) a cross-cutting
comparative analysis and discussion of open challenges.

The remainder of this paper is organized as follows. Section~\ref{sec:past}
recapitulates past work. Section~\ref{sec:gan} covers GAN-based methods.
Section~\ref{sec:diffusion} discusses diffusion-based approaches.
Section~\ref{sec:nerf} reviews neural radiance field methods.
Section~\ref{sec:cnn} presents complementary CNN approaches.
Section~\ref{sec:comparison} provides comparative analysis.
Section~\ref{sec:challenges} discusses challenges and future directions.
Section~\ref{sec:conclusion} concludes the paper.

%% =========================================================
\section{Recapitulation of Past Work: From Statistical Models to Early CNNs}
\label{sec:past}
%% =========================================================

Before examining the modern generative era, it is essential to situate it
within the trajectory of earlier methods. The foundational review by Maken and
Gupta~\cite{maken2023} provides the most comprehensive taxonomy of pre-2020
approaches across six categories: intensity-based methods, statistical shape
models, articulated anatomical models, free-form and hierarchical deformation,
convolutional neural networks, and hybrid methods. Figure~\ref{fig:general_pipeline}
illustrates the general pipeline for 3D reconstruction from X-ray images.

\begin{figure}[h]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[node distance=0.35cm and 0.2cm]
  \node[block,text width=1.3cm,minimum height=0.85cm] (acq)
    {X-ray\\ Acq.\\ (1--2 views)};
  \node[block,text width=1.3cm,minimum height=0.85cm,right=of acq] (enh)
    {Image\\ Enhance-\\ ment};
  \node[block,text width=1.3cm,minimum height=0.85cm,right=of enh] (seg)
    {Contour\\ Extract.\\ (Seg.)};
  \node[block,text width=1.3cm,minimum height=0.85cm,right=of seg] (reg)
    {2D/3D\\ Corresp.\\ \& Reg.};
  \node[blockG,text width=1.3cm,minimum height=0.85cm,right=of reg] (rec)
    {3D\\ Recon-\\ struction};
  \node[blockG,text width=1.2cm,minimum height=0.85cm,right=of rec] (out)
    {3D Vol.\\ / Mesh\\ Output};
  \foreach \a/\b in {acq/enh,enh/seg,seg/reg,reg/rec,rec/out}
    {\draw[arrow] (\a) -- (\b);}
\end{tikzpicture}}
\caption{General pipeline for 3D anatomical reconstruction from 2D X-ray
images applicable across all method categories~\cite{maken2023}.}
\label{fig:general_pipeline}
\end{figure}

\subsection{Statistical Shape Models (SSMs) and Parametric Approaches}
\label{sec:past:ssm}

Statistical Shape Models, pioneered by Cootes et al.~\cite{cootes1992} through
Point Distribution Models (PDMs), represent anatomical shape as a mean shape
plus a weighted combination of principal components of shape variation:
\begin{equation}
\mathbf{S}' = M\bar{S} + M\sum_{j=1}^{s} f(\lambda_j)\mathbf{P}_j,\quad
f(\lambda_j) = a_j = \sqrt{\lambda_j}
\label{eq:ssm}
\end{equation}
where $\bar{S}$ is the population mean shape, $\mathbf{P}_j$ are eigenvectors,
$\lambda_j$ are eigenvalues from PCA, and $a_j$ are weight vectors. The
covariance matrix is:
\begin{equation}
\text{Cov} = \frac{1}{p-1}\sum_{i=1}^{p}(S_i - \bar{S})(S_i - \bar{S})^T
\label{eq:cov}
\end{equation}
Figure~\ref{fig:ssm_pipeline} illustrates the SSM-based reconstruction pipeline.

\begin{figure}[h]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[node distance=0.3cm and 0.2cm]
  \node[blockO,text width=1.4cm,minimum height=0.75cm] (ct) {CT Training DB};
  \node[blockO,text width=1.4cm,minimum height=0.75cm,right=of ct] (seg2)
    {Seg. \& Land-\\ marking};
  \node[blockO,text width=1.3cm,minimum height=0.75cm,right=of seg2] (pca)
    {PCA\\ (Eq.~\ref{eq:cov})};
  \node[blockO,text width=1.3cm,minimum height=0.75cm,right=of pca] (ssm)
    {SSM\\ Build};
  \foreach \a/\b in {ct/seg2,seg2/pca,pca/ssm}{\draw[arrow] (\a) -- (\b);}
  \node[above=0.1cm of ct,font=\tiny\bfseries,orange!80!black]
    {\textit{Training Phase}};

  \node[block,text width=1.4cm,minimum height=0.75cm,below=0.8cm of ct] (xray)
    {Patient X-ray};
  \node[block,text width=1.3cm,minimum height=0.75cm,right=of xray] (cont)
    {Contour Extract.};
  \node[block,text width=1.3cm,minimum height=0.75cm,right=of cont] (fit)
    {SSM Fit\\ (opt. $a_j$)};
  \node[blockG,text width=1.3cm,minimum height=0.75cm,right=of fit] (vol)
    {3D Surface};
  \foreach \a/\b in {xray/cont,cont/fit,fit/vol}{\draw[arrow] (\a) -- (\b);}
  \node[below=0.1cm of xray,font=\tiny\bfseries,blue!70!black]
    {\textit{Inference Phase}};
  \draw[arrow,dashed,orange!60] (ssm.south) -- ++(0,-0.25) -| (fit.north);
\end{tikzpicture}}
\caption{SSM-based reconstruction pipeline. An SSM is built offline from CT
training data via PCA. At inference time, SSM parameters are optimized to match
X-ray contours, yielding a patient-specific 3D surface model~\cite{lamecker2006,zheng2010}.}
\label{fig:ssm_pipeline}
\end{figure}

Lamecker et al.~\cite{lamecker2006} demonstrated SSM-based reconstruction of
the pelvic bone from synthetically generated X-rays, achieving symmetric mean
errors of approximately 1.3--2.6\,mm. Zheng~\cite{zheng2010} extended this to
single-view AP X-ray reconstruction of the pelvis using a hybrid 2D/3D
deformable registration strategy. Zhu and Li~\cite{zhu2011} applied enhanced
SSM with 2D/3D registration to the distal femur, reporting an overall average
absolute error of 0.90\,mm. Parametric model-based reconstruction (Humbert et
al.~\cite{humbert2009} for spine; Chaibi et al.~\cite{chaibi2012} for lower
limb) adjusts anatomically meaningful parameters to fit radiographic
projections, achieving reconstruction times of 2--10\,minutes.

\subsection{Template Deformation Methods}
\label{sec:past:deform}

Template deformation approaches deform a canonical 3D mesh model until its
projected silhouette matches X-ray contours. Free-form deformation (FFD) was
demonstrated by Koh et al.~\cite{koh2011} for femur reconstruction. Hierarchical
FFD by Gunay et al.~\cite{gunay2007} applies multi-resolution deformation for
surgical osteotomy planning. Karade and Ravi~\cite{karade2015} introduced
Laplacian Surface Deformation (LSD) with Self-Organizing Maps (SOM) for 2D/3D
correspondence, achieving mean point-to-surface distance of 1.2\,mm in under
49\,seconds.

\subsection{Early CNN-Based Methods}
\label{sec:past:cnn}

The transition to data-driven approaches began with CNNs. Dixit et
al.~\cite{dixit2019} proposed a feed-forward CNN to reconstruct 3D femur models
from 30 X-ray samples. Kim et al.~\cite{kim2019} combined CNN-based feature
detection (YOLO v3 and Faster R-CNN) with SSM fitting for leg bone
reconstruction, achieving 187-second computation time. The landmark contribution
from Kasten et al.~\cite{kasten2020} introduced an end-to-end CNN for 3D knee
bone reconstruction from biplanar X-rays using dimensional augmentation that
back-projects epipolar line pairs into a two-channeled epipolar plane, achieving
a Dice coefficient of 0.906 with 0.5-second inference---a step change in
reconstruction speed.

%% =========================================================
\section{GAN-Based Methods for X-ray to CT Reconstruction}
\label{sec:gan}
%% =========================================================

Generative Adversarial Networks (GANs)~\cite{goodfellow2014} consist of a
generator $G$ trained to synthesize realistic outputs and a discriminator $D$
trained to distinguish real from generated samples. In medical image synthesis,
adversarial training produces volumetric CT outputs with sharper detail and more
realistic texture than pure regression-based methods. Figure~\ref{fig:gan_overview}
illustrates the general GAN architecture for X-ray to CT synthesis.


\begin{figure*}[t]
\centering
\begin{tikzpicture}[node distance=0.45cm and 0.45cm,
    every node/.style={font=\scriptsize}]
  %% INPUTS
  \node[block,text width=1.3cm,minimum height=1.1cm] (ap) {AP\\ X-ray};
  \node[block,text width=1.3cm,minimum height=1.1cm,below=0.3cm of ap] (lat)
    {LAT\\ X-ray};
  %% 2D ENCODERS
  \node[blockO,text width=1.5cm,minimum height=1.1cm,right=0.5cm of ap] (encA)
    {2D Encoder\\ Branch A};
  \node[blockO,text width=1.5cm,minimum height=1.1cm,right=0.5cm of lat] (encB)
    {2D Encoder\\ Branch B};
  \draw[arrow] (ap) -- (encA);
  \draw[arrow] (lat) -- (encB);
  %% FUSION
  \node[blockO,text width=1.5cm,minimum height=2.7cm,
    right=0.5cm of encA,yshift=-0.7cm] (fuse)
    {Feature\\ Fusion\\ (Concat /\\ Attention)};
  \draw[arrow] (encA) -- (fuse);
  \draw[arrow] (encB) -- (fuse);
  %% GENERATOR
  \node[blockG,text width=1.8cm,minimum height=2.7cm,
    right=0.5cm of fuse] (gen)
    {\textbf{Generator}\\ G\\ 3D Decoder\\ (Conv3D /\\ Transformer)};
  \draw[arrow] (fuse) -- (gen);
  %% OUTPUTS
  \node[blockG,text width=1.3cm,minimum height=1.1cm,
    right=0.5cm of gen,yshift=0.8cm] (fake) {Synth. CT\\ $\hat{V}$};
  \node[block,text width=1.3cm,minimum height=1.1cm,
    right=0.5cm of gen,yshift=-0.8cm] (real) {Real CT\\ $V$};
  \draw[arrow] (gen) -- (fake);
  %% DISCRIMINATOR
  \node[blockR,text width=1.8cm,minimum height=2.7cm,
    right=0.5cm of fake,yshift=-0.8cm] (disc)
    {\textbf{Discriminator}\\ D\\ 3D-PatchGAN\\ Real / Fake};
  \draw[arrow] (fake) -- (disc);
  \draw[arrow] (real) -- (disc);
  %% LOSSES
  \node[blockR,text width=1.6cm,minimum height=0.75cm,
    right=0.5cm of disc,yshift=0.9cm] (ladv) {$\mathcal{L}_\text{adv}$};
  \node[blockP,text width=1.6cm,minimum height=0.75cm,below=0.2cm of ladv] (lperc)
    {$\mathcal{L}_\text{perceptual}$};
  \node[blockO,text width=1.6cm,minimum height=0.75cm,below=0.2cm of lperc] (ll1)
    {$\mathcal{L}_{L_1}$};
  \node[blockR,text width=2.0cm,minimum height=0.75cm,below=0.25cm of ll1] (ltot)
    {$\mathcal{L}_\text{total}$};
  \draw[arrow] (disc) -- (ladv);
  \draw[arrow] (ladv.south) -- (lperc.north);
  \draw[arrow] (lperc.south) -- (ll1.north);
  \draw[arrow] (ll1.south) -- (ltot.north);
  \draw[arrow,dashed,draw=red!50] (ladv.west) -- ++(-0.25,0) |- (gen.east)
    node[pos=0.2,right,font=\tiny,red!50] {update G};
  %% DRR consistency
  \node[blockP,text width=1.5cm,minimum height=0.7cm,below=0.5cm of gen] (drr)
    {DRR Proj.\\ Consistency};
  \draw[arrow,dashed] (gen.south) -- (drr.north);
  \draw[arrow,dashed] (drr.east) -| (lperc.south);
  %% Background boxes
  \begin{scope}[on background layer]
    \node[draw=orange!40,fill=orange!5,rounded corners,
      inner sep=4pt,fit=(encA)(encB)(fuse)(gen),
      label={[font=\small\bfseries,orange!80!black]above:Generator (G)}] {};
    \node[draw=red!40,fill=red!5,rounded corners,
      inner sep=4pt,fit=(disc)(ladv)(lperc)(ll1)(ltot),
      label={[font=\small\bfseries,red!70]above:Discriminator (D)}] {};
  \end{scope}
\end{tikzpicture}
\caption{General GAN-based architecture for X-ray to CT reconstruction.
Biplanar AP and LAT X-rays are independently encoded, fused, and decoded by a
3D generator. The discriminator distinguishes real from generated volumes.
Multiple loss terms ($\mathcal{L}_\text{adv}$, $\mathcal{L}_\text{perceptual}$,
$\mathcal{L}_{L_1}$) jointly optimize the generator~\cite{x2ctgan,trctgan}.}
\label{fig:gan_overview}
\end{figure*}

\subsection{Foundational Architecture: X2CT-GAN}
\label{sec:gan:x2ct}

X2CT-GAN established the canonical biplanar X-ray input paradigm for
GAN-based volumetric synthesis. Taking orthogonal anteroposterior (AP) and
lateral (LAT) chest X-ray images as input, it generates a full 3D chest CT
volume via two separate 2D-to-3D encoding branches followed by feature fusion
and a 3D decoder. The discriminator operates on both 2D projections of the
generated volume and the 3D volume itself, exploiting complementary depth
information from perpendicular views.

\subsection{CVAE-GAN: Probabilistic 3D CT Reconstruction}
\label{sec:gan:cvae}

Jiang et al.~\cite{cvaegan} proposed a Conditional Variational Autoencoder
combined with a GAN discriminator (CVAE-GAN) for 3D CT reconstruction from a
single X-ray, specifically targeting respiratory motion management in
radiotherapy. The encoder maps a 2D X-ray projection into a latent distribution
$(\mu, \sigma^2)$, from which a latent vector $\mathbf{z} \sim \mathcal{N}(\mu,
\sigma^2)$ is sampled and decoded into a 3D CT volume. The training objective
is:
\begin{equation}
\mathcal{L} = \mathcal{L}_\text{recon} + \beta\,D_\text{KL}
\!\bigl(\mathcal{N}(\mu,\sigma^2)\,\|\,\mathcal{N}(0,I)\bigr)
+ \mathcal{L}_\text{adv}
\label{eq:cvae}
\end{equation}
By sampling from a learned latent distribution, the CVAE component enables
representing the inherent ambiguity in reconstructing 3D anatomy from a single
2D projection, providing a clinically valuable implicit uncertainty estimate.

\subsection{TRCT-GAN: Transformer-Enhanced Biplanar Reconstruction}
\label{sec:gan:trct}

Wang et al.~\cite{trctgan} proposed TRCT-GAN, which enhances biplanar X-ray
feature representation through transformer modules. While convolutional
operations capture local spatial patterns efficiently, they struggle to model
long-range dependencies such as the relationship between the left and right lung
fields. TRCT-GAN incorporates a Transformer module within the generator to
capture global contextual relationships, and a dynamic attention module to
exploit both 2D and 3D feature maps. Figure~\ref{fig:trct} illustrates the
dynamic attention mechanism.

\begin{figure}[h]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[node distance=0.35cm and 0.3cm]
  \node[block,text width=1.3cm,minimum height=0.8cm] (xA) {AP X-ray};
  \node[block,text width=1.3cm,minimum height=0.8cm,below=0.3cm of xA] (xL)
    {LAT X-ray};
  \node[blockP,text width=1.6cm,minimum height=2.0cm,
    right=0.4cm of xA,yshift=-0.55cm] (trans)
    {Transformer\\ Multi-Head\\ Self-Attention};
  \draw[arrow] (xA) -- (trans);
  \draw[arrow] (xL) -- (trans);
  \node[blockO,text width=1.5cm,minimum height=0.8cm,
    right=0.4cm of trans,yshift=0.6cm] (f2d) {2D Feature\\ $F_{2D}$};
  \node[blockG,text width=1.5cm,minimum height=0.8cm,
    right=0.4cm of trans,yshift=-0.6cm] (f3d) {3D Feature\\ $F_{3D}$};
  \draw[arrow] (trans) -- (f2d);
  \draw[arrow] (trans) -- (f3d);
  \node[blockR,text width=1.7cm,minimum height=2.0cm,
    right=0.4cm of f2d,yshift=-0.45cm] (dyn)
    {Dynamic\\ Attention\\ Module\\ (Cross-Domain)};
  \draw[arrow] (f2d) -- (dyn);
  \draw[arrow] (f3d) -- (dyn);
  \node[blockG,text width=1.3cm,minimum height=0.8cm,right=0.4cm of dyn]
    (out) {CT Volume};
  \draw[arrow] (dyn) -- (out);
\end{tikzpicture}}
\caption{TRCT-GAN architecture~\cite{trctgan}. A Transformer module captures
long-range X-ray dependencies. A Dynamic Attention Module bridges 2D input
features and 3D volumetric output features through cross-domain contextual
association.}
\label{fig:trct}
\end{figure}

\subsection{DP-GAN+B: Lightweight GAN with Depthwise Separable Convolutions}
\label{sec:gan:dpgan}

Xing et al.~\cite{dpganb} introduced DP-GAN+B to address the computational
efficiency bottleneck limiting clinical deployment. DP-GAN+B replaces standard
convolutions with depthwise separable convolutions:
\begin{equation}
\text{Params}_\text{DW-Sep} = D_K^2 C_\text{in} + C_\text{in} C_\text{out}
\ll D_K^2 C_\text{in} C_\text{out} = \text{Params}_\text{Std}
\label{eq:dwsep}
\end{equation}
achieving a parameter reduction of 31.924 million (44.17\%) compared to
X2CT-GAN baseline, while also introducing vector loss and fusion loss components.

\subsection{AP2CT-GAN: Single AP View Reconstruction for ICU Settings}
\label{sec:gan:ap2ct}

Tang et al.~\cite{ap2ctgan} proposed AP2CT-GAN for the ICU clinical scenario
where only a single anteroposterior (AP) chest X-ray can practically be
acquired. Two key modules are introduced: the Feature Enhancement Connection
(FEC) module amplifying anatomically informative regions, and the Feature
Dimension Converter (FDC) module bridging 2D and 3D domains. A Dual-Consistency
Loss enforces structural and textural consistency:
\begin{equation}
\mathcal{L}_\text{DC} = \alpha\,\mathcal{L}_\text{struct}
+ (1-\alpha)\,\mathcal{L}_\text{texture}
\label{eq:dcloss}
\end{equation}

\subsection{PTX2CT-GAN: Disease-Specific GAN for Pneumothorax}
\label{sec:gan:ptx}

Wang et al.~\cite{ptx2ctgan} presented PTX2CT-GAN targeting reconstruction of
3D chest CT from a single X-ray in the context of pneumothorax (PTX). The
Reconstruction Feature Enhancement Module (RFEM) improves feature extraction
specifically for pathological pneumothorax features. Trained and validated on
the MY-PTX-998 dataset (998 pneumothorax chest X-rays with paired CT),
PTX2CT-GAN demonstrates the value of disease-specific architectural design.

Table~\ref{tab:gan_comparison} summarizes all GAN-based methods reviewed.

\begin{table*}[t]
\caption{Comparison of GAN-Based X-ray to CT Reconstruction Methods}
\label{tab:gan_comparison}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}llllp{2.6cm}p{2.0cm}p{2.8cm}@{}}
\toprule
\textbf{Method} & \textbf{Year} & \textbf{Input} & \textbf{Region} &
\textbf{Key Innovation} & \textbf{Dataset} & \textbf{Performance}\\
\midrule
CVAE-GAN~\cite{cvaegan} & 2021 & Single X-ray & Lung &
  Probabilistic VAE + GAN; uncertainty estimation & 2 patient CTs &
  Feasibility; qualitative assessment\\
X-CTRSNet~\cite{xctrsnet} & 2022 & Biplanar AP+LAT & C-Vertebra &
  Joint recon.\ + segmentation (RSC loss) & Institutional CT &
  PSNR=24.58\,dB, SSIM=0.749, Dice=80.44\%\\
XctNet~\cite{xctnet} & 2022 & Single X-ray & General &
  Self-attn.\ + multi-scale + 3D branch gen. & Public dataset &
  PSNR=29.28\,dB, SSIM=0.8681\\
TRCT-GAN~\cite{trctgan} & 2023 & Biplanar & Chest CT &
  Transformer + dynamic cross-domain attention & ChestX-ray14/LIDC &
  Improved SSIM/PSNR vs X2CT-GAN\\
DP-GAN+B~\cite{dpganb} & 2024 & Biplanar & Lung CT &
  Depthwise sep.\ convs.; 44.17\% param reduction & Institutional CT &
  Comparable quality; $-31.9$M params\\
AP2CT-GAN~\cite{ap2ctgan} & 2025 & Single AP & Chest CT &
  FEC + FDC + dual-consistency loss & Institutional &
  Best single-view metrics vs baselines\\
PTX2CT-GAN~\cite{ptx2ctgan} & 2025 & Single X-ray & Pneumothorax &
  RFEM for disease-specific features & MY-PTX-998 &
  Superior PTX-specific quantitative metrics\\
\bottomrule
\end{tabular}
\end{table*}

%% =========================================================
\section{Diffusion Models for X-ray to CT Reconstruction}
\label{sec:diffusion}
%% =========================================================

Diffusion models (DDPMs)~\cite{ho2020ddpm} operate by learning to reverse a
gradual noising process. During training, Gaussian noise is progressively added
to clean data over $T$ timesteps:
\begin{equation}
q(\mathbf{x}_t|\mathbf{x}_{t-1}) =
\mathcal{N}\!\bigl(\mathbf{x}_t;\sqrt{1-\beta_t}\mathbf{x}_{t-1},\beta_t\mathbf{I}\bigr)
\label{eq:forward}
\end{equation}
The model learns to reverse this process, conditioned on X-ray images
$\mathbf{c}$:
\begin{equation}
p_\theta(\mathbf{x}_{t-1}|\mathbf{x}_t, \mathbf{c}) =
\mathcal{N}\!\bigl(\mathbf{x}_{t-1};\mu_\theta(\mathbf{x}_t,t,\mathbf{c}),\Sigma_\theta\bigr)
\label{eq:reverse}
\end{equation}
Diffusion models offer several advantages over GANs: training is stable (no
adversarial game), mode coverage is better, and iterative refinement allows
progressive improvement of reconstruction quality. Figure~\ref{fig:diffusion_overview}
illustrates the diffusion-based reconstruction framework.

\begin{figure*}[t]
\centering
\begin{tikzpicture}[node distance=0.45cm and 0.4cm,
    every node/.style={font=\scriptsize}]
  %% FORWARD PROCESS (top row)
  \node[blockG,text width=1.2cm,minimum height=0.9cm] (ct0)
    {CT Vol.\\ $\mathbf{x}_0$};
  \node[blockO,text width=1.2cm,minimum height=0.9cm,right=0.45cm of ct0]
    (ct1) {$\mathbf{x}_1$\\ (slight\\ noise)};
  \node[blockO,text width=1.2cm,minimum height=0.9cm,right=0.45cm of ct1]
    (ct2) {$\mathbf{x}_t$\\ (partial\\ noise)};
  \node[blockR,text width=1.2cm,minimum height=0.9cm,right=0.45cm of ct2]
    (ctT) {$\mathbf{x}_T$\\ (pure\\ Gaussian)};
  \draw[arrow] (ct0) -- node[above,font=\tiny]{$q(\mathbf{x}_t|\mathbf{x}_{t-1})$} (ct1);
  \draw[arrow,dashed] (ct1) -- (ct2);
  \draw[arrow,dashed] (ct2) -- (ctT);
  \node[above=0.1cm of ct0,orange!80!black,font=\scriptsize\bfseries]
    {Forward Process (noising) $\longrightarrow$};

  %% REVERSE PROCESS (bottom row)
  \node[blockR,text width=1.2cm,minimum height=0.9cm,
    below=1.4cm of ctT] (rT) {$\mathbf{x}_T$\\ noise};
  \node[blockP,text width=1.2cm,minimum height=0.9cm,left=0.45cm of rT]
    (rt) {$\mathbf{x}_t$\\ denoise};
  \node[blockP,text width=1.2cm,minimum height=0.9cm,left=0.45cm of rt]
    (r1) {$\mathbf{x}_1$\\ almost\\ clean};
  \node[blockG,text width=1.2cm,minimum height=0.9cm,left=0.45cm of r1]
    (r0) {$\hat{\mathbf{x}}_0$\\ synth.\ CT};
  \draw[arrow] (rT) -- node[above,font=\tiny]
    {$p_\theta(\mathbf{x}_{t-1}|\mathbf{x}_t,\mathbf{c})$} (rt);
  \draw[arrow,dashed] (rt) -- (r1);
  \draw[arrow,dashed] (r1) -- (r0);
  \node[below=0.1cm of rT,purple!80!black,font=\scriptsize\bfseries]
    {$\longleftarrow$ Reverse Process (denoising, conditioned on X-ray)};

  %% CONDITIONING
  \node[block,text width=1.8cm,minimum height=0.9cm,
    below=0.4cm of ct2] (xraya) {X-ray(s)\\ (AP / LAT)};
  \node[blockP,text width=2.0cm,minimum height=0.9cm,
    right=0.4cm of xraya] (enc2)
    {Position-Aware\\ X-ray Encoder\\ (Transformer)};
  \draw[arrow] (xraya) -- (enc2);
  \draw[arrow,dashed,purple!60] (enc2.north)
    -- ++(0,0.25) -| (rt.south)
    node[pos=0.1,right,font=\tiny,purple!60]{cond.\ $\mathbf{c}$};

  %% Denoising network label
  \node[blockP,text width=1.7cm,minimum height=0.8cm,
    above=0.3cm of rt] (dnet) {Denoising Net\\ $\epsilon_\theta(\mathbf{x}_t,t,\mathbf{c})$};
  \draw[darrow,gray!40] (rt) -- (dnet);
\end{tikzpicture}
\caption{Diffusion model framework for X-ray to CT reconstruction. The forward
process progressively adds Gaussian noise ($\mathbf{x}_0 \rightarrow
\mathbf{x}_T$, Eq.~\ref{eq:forward}). The conditioned reverse process
denoises from pure Gaussian noise to produce synthetic CT
$\hat{\mathbf{x}}_0$~\cite{dx2ct,dvgdiffusion,yoon2024}.}
\label{fig:diffusion_overview}
\end{figure*}

\subsection{DX2CT: Positional-Aware Diffusion for Bi/Mono-Planar Reconstruction}
\label{sec:diffusion:dx2ct}

Jeong et al.~\cite{dx2ct} proposed DX2CT, a conditional diffusion model for 3D
CT reconstruction from either biplanar or monoplanar X-ray images. DX2CT
resolves the positional ambiguity of prior GAN-based methods through two key
innovations. First, a novel transformer modulates 2D X-ray feature maps with 3D
positional information of the CT volume via positional embeddings:
\begin{equation}
\mathbf{F}_{3D}^{pos}(i,j,k) = \text{CrossAttn}\bigl(
\mathbf{E}_{pos}(i,j,k),\,\mathbf{F}_{2D}\bigr)
\label{eq:dx2ct}
\end{equation}
where $\mathbf{E}_{pos}(i,j,k)$ are 3D positional encodings of CT voxel
$(i,j,k)$ querying relevant 2D X-ray features $\mathbf{F}_{2D}$ in a
geometrically principled manner. Second, these position-aware features condition
the diffusion denoising network at each reverse step. Experiments on both
biplanar and monoplanar X-ray benchmark datasets demonstrate that DX2CT
outperforms several state-of-the-art methods.

\subsection{DVG-Diffusion: Dual-View Guided Latent Diffusion}
\label{sec:diffusion:dvg}

Xie et al.~\cite{dvgdiffusion} proposed DVG-Diffusion, a dual-view guided
diffusion model that directly addresses the ill-posed nature of mapping few-view
2D X-rays to 3D CT. The core insight is that a synthetic additional view can be
generated and used alongside the real view to jointly guide reconstruction.
DVG-Diffusion consists of three components: (1)~a view-parameter-guided encoder
extracting geometrically consistent features from real X-ray images;
(2)~a novel view synthesis module generating a synthetic X-ray from a different
angle; and (3)~a latent diffusion model conditioned on dual-view features,
operating in the compressed latent space of a pre-trained VAE. Figure~\ref{fig:dvg}
illustrates this architecture.

\begin{figure}[h]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[node distance=0.35cm and 0.3cm]
  \node[block,text width=1.4cm,minimum height=0.75cm] (realX)
    {Real X-ray\\ (view $\phi_1$)};
  \node[blockO,text width=1.5cm,minimum height=0.75cm,right=0.35cm of realX]
    (venc) {View-Param\\ Guided Enc.};
  \draw[arrow] (realX) -- (venc);
  \node[blockO,text width=1.4cm,minimum height=0.65cm,above=0.25cm of venc]
    (vpar) {View Params\\ $(\phi_1, d_{sd})$};
  \draw[arrow] (vpar) -- (venc);
  \node[blockP,text width=1.5cm,minimum height=0.75cm,below=0.25cm of venc]
    (vsyn) {View Synth.\\ Module\\ ($\hat{\phi}_2$)};
  \draw[arrow] (venc) -- (vsyn);
  \node[blockG,text width=1.4cm,minimum height=1.6cm,
    right=0.35cm of venc,yshift=-0.35cm] (cat) {Dual-View\\ Feature\\ Concat};
  \draw[arrow] (venc) -- (cat);
  \draw[arrow] (vsyn) -- (cat);
  \node[blockP,text width=1.4cm,minimum height=0.75cm,
    right=0.35cm of cat,yshift=0.35cm] (vae) {VAE Latent\\ Encoder};
  \draw[arrow] (cat) -- (vae);
  \node[blockR,text width=1.4cm,minimum height=0.75cm,right=0.35cm of vae]
    (ldm) {Latent Diff.\\ Model};
  \draw[arrow] (vae) -- (ldm);
  \node[blockG,text width=1.4cm,minimum height=0.75cm,right=0.35cm of ldm]
    (vdec) {VAE Dec.\\ $\hat{V}$};
  \draw[arrow] (ldm) -- (vdec);
\end{tikzpicture}}
\caption{DVG-Diffusion architecture~\cite{dvgdiffusion}. A view-parameter-guided
encoder and view synthesis module generate dual-view features conditioning a
Latent Diffusion Model (LDM) operating in VAE latent space.}
\label{fig:dvg}
\end{figure}

\subsection{High-Resolution 3D CT Synthesis Using Patch-Based Diffusion}
\label{sec:diffusion:yoon}

Yoon et al.~\cite{yoon2024} proposed a 3D diffusion model for synthesizing
high-resolution CT volumes from bidirectional X-ray images, targeting orthopedic
applications (wrist fractures). To address computational cost, the 3D diffusion
model is trained using patch-wise loss---operating on volumetric patches rather
than the full CT volume, enabling single-GPU training. Patch coordinate
information compensates for limited training examples by providing spatial
context within the conditioning signal.

Table~\ref{tab:diffusion_comparison} summarizes diffusion-based methods.

\begin{table*}[t]
\caption{Comparison of Diffusion-Based X-ray to CT Reconstruction Methods}
\label{tab:diffusion_comparison}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}llllp{2.8cm}p{2.0cm}p{3.0cm}@{}}
\toprule
\textbf{Method} & \textbf{Year} & \textbf{Input} & \textbf{Region} &
\textbf{Key Innovation} & \textbf{Dataset} & \textbf{Key Result}\\
\midrule
3D Diffusion~\cite{yoon2024} & 2024 & Bidirectional (2) & Wrist/Bone CT &
  Patch-wise loss; single-GPU; high-res synthesis & Institutional wrist &
  High-fidelity 3D CT; single GPU feasible\\
DX2CT~\cite{dx2ct} & 2025 & Mono or biplanar & Chest CT &
  3D position-aware transformer conditioning & LIDC-IDRI &
  SOTA on bi- and mono-planar benchmarks\\
DVG-Diffusion~\cite{dvgdiffusion} & 2026 & Few-view & Chest/General &
  View synthesis + dual-view latent diffusion & ChestX-ray14/LIDC &
  Best fidelity-perceptual balance; SOTA IEEE TIP\\
\bottomrule
\end{tabular}
\end{table*}

%% =========================================================
\section{Neural Radiance Fields for Medical 3D Reconstruction}
\label{sec:nerf}
%% =========================================================

Neural Radiance Fields (NeRF)~\cite{nerf2020} learn a continuous implicit
function mapping a 3D position and viewing direction to density and radiance.
In medical imaging, CT volumes can be interpreted as 3D density fields, and
X-ray projections as rendered views. New views are synthesized via volumetric
rendering:
\begin{equation}
C(\mathbf{r}) = \int_{t_n}^{t_f} T(t)\,\sigma(\mathbf{r}(t))\,
\mathbf{c}(\mathbf{r}(t),\mathbf{d})\,dt, \quad
T(t) = e^{-\int_{t_n}^{t}\sigma(\mathbf{r}(s))\,ds}
\label{eq:nerf}
\end{equation}
Figure~\ref{fig:nerf_arch} illustrates the NeRF framework adapted for medical
imaging.

\begin{figure*}[t]
\centering
\begin{tikzpicture}[node distance=0.45cm and 0.5cm,
    every node/.style={font=\scriptsize}]
  \node[block,text width=1.3cm,minimum height=0.9cm] (xray) {Single\\ X-ray};
  \node[blockO,text width=1.7cm,minimum height=0.9cm,right=0.5cm of xray]
    (enc) {2D CNN Encoder\\ (ConvNeXt /\\ ResNet)};
  \draw[arrow] (xray) -- (enc);
  \node[blockO,text width=1.4cm,minimum height=0.9cm,right=0.5cm of enc]
    (latent) {Latent Code\\ $\mathbf{z}$};
  \draw[arrow] (enc) -- (latent);
  \node[blockP,text width=1.4cm,minimum height=0.9cm,below=0.9cm of latent]
    (pos) {3D Query\\ $(x,y,z)$\\ direction $\mathbf{d}$};
  \node[blockP,text width=2.0cm,minimum height=1.8cm,
    right=0.5cm of latent,yshift=-0.45cm] (mlp)
    {\textbf{NeRF MLP}\\
    $F_\theta(x,y,z,\mathbf{d}\,|\,\mathbf{z})$\\[3pt]
    $\downarrow$\\ $(\sigma,\mathbf{c})$};
  \draw[arrow] (latent) -- (mlp);
  \draw[arrow] (pos) -- (mlp);
  \node[blockG,text width=1.7cm,minimum height=0.9cm,
    right=0.5cm of mlp] (render) {Volumetric\\ Rendering\\ (Eq.~\ref{eq:nerf})};
  \draw[arrow] (mlp) -- (render);
  \node[blockG,text width=1.4cm,minimum height=0.9cm,
    right=0.5cm of render,yshift=0.5cm] (ctout) {Synth.\ CT\\ Projection\\ $\hat{C}(\mathbf{r})$};
  \node[blockG,text width=1.4cm,minimum height=0.9cm,
    right=0.5cm of render,yshift=-0.5cm] (vol3d) {3D Density\\ Volume\\ $\sigma(x,y,z)$};
  \draw[arrow] (render) -- (ctout);
  \draw[arrow] (render) -- (vol3d);
  \node[blockR,text width=1.7cm,minimum height=0.8cm,below=0.5cm of render]
    (loss) {Rendering Loss\\ $\|\hat{C}-C_\text{GT}\|^2$};
  \draw[arrow,dashed] (ctout.south) |- (loss.east);
  \draw[arrow,dashed,red!50] (loss.west) -| (mlp.south)
    node[pos=0.3,below,font=\tiny,red!50]{backprop};
  \node[blockP,text width=2.0cm,minimum height=0.75cm,below=0.3cm of mlp]
    (disent) {Shape / Depth\\ Disentanglement\\ (MedNeRF)};
  \draw[arrow,dashed] (mlp.south) -- (disent.north);
\end{tikzpicture}
\caption{Neural Radiance Field (NeRF) architecture adapted for medical imaging.
A 2D CNN encoder extracts latent code $\mathbf{z}$ from a single X-ray. The
NeRF MLP maps 3D coordinates conditioned on $\mathbf{z}$ to density $\sigma$
and radiance $\mathbf{c}$. Volumetric rendering synthesizes CT projections from
arbitrary angles, and the 3D density volume constitutes the reconstructed
CT~\cite{mednerf,x2br}.}
\label{fig:nerf_arch}
\end{figure*}

\subsection{MedNeRF: Medical Neural Radiance Fields}
\label{sec:nerf:mednerf}

Corona-Figueroa et al.~\cite{mednerf} proposed MedNeRF, the first application of
neural radiance fields to the problem of reconstructing 3D-aware CT projections
from a single or few X-ray images. MedNeRF introduces a novel design that
disentangles shape and volumetric depth, learning a continuous representation of
CT scans by separately modelling surface geometry and internal density from 2D
X-ray observations. Validated on chest and knee datasets, MedNeRF demonstrates
that a single input X-ray is sufficient to synthesize new CT projections from
arbitrary viewing angles. By reducing the number of X-ray projections required
to infer 3D anatomy from hundreds (CT) to one, it has the potential to
substantially reduce patient radiation exposure.

\subsection{X2BR: Hybrid Neural Implicit Framework for Bone Reconstruction}
\label{sec:nerf:x2br}

Guven et al.~\cite{x2br} proposed X2BR, combining continuous volumetric
reconstruction with template-guided non-rigid registration for high-fidelity 3D
bone reconstruction from a single planar X-ray. X2BR comprises two components.
The first, X2B, employs a ConvNeXt-based encoder to predict a 3D bone occupancy
field using an implicit neural representation:
\begin{equation}
f_\theta: (\mathbf{x},\mathbf{z}) \mapsto o \in [0,1]
\label{eq:x2b}
\end{equation}
where $\mathbf{x} \in \mathbb{R}^3$ is a 3D query point and $\mathbf{z}$ is
the X-ray latent code. X2B achieves IoU of 0.952 and Chamfer-L1 distance of
0.005, without relying on statistical shape models. The second component
integrates anatomical priors via YOLOv9-based landmark detection and
geodesic-based coherent point drift (CPD) registration, improving anatomical
consistency (IoU: 0.875 for X2BR).

Table~\ref{tab:nerf_comparison} summarizes NeRF and neural implicit methods.

\begin{table*}[t]
\caption{Comparison of Neural Radiance Field and Neural Implicit Methods}
\label{tab:nerf_comparison}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}llllp{2.5cm}p{2.0cm}p{3.2cm}@{}}
\toprule
\textbf{Method} & \textbf{Year} & \textbf{Input} & \textbf{Region} &
\textbf{Representation} & \textbf{Dataset} & \textbf{Performance}\\
\midrule
MedNeRF~\cite{mednerf} & 2022 & Single / few X-rays & Chest / Knee &
  NeRF + shape--depth disentanglement & Chest \& knee CTs &
  High-fidelity CT projection synthesis; SOTA NeRF comparison\\
X2BR~\cite{x2br} & 2026 & Single planar X-ray & Bone (Rib, Vertebra) &
  ConvNeXt + neural occupancy + CPD registration & Institutional bone CT &
  X2B: IoU=0.952, Chamfer=0.005; X2BR: superior anatomical realism\\
\bottomrule
\end{tabular}
\end{table*}

%% =========================================================
\section{Complementary CNN and Transformer Approaches}
\label{sec:cnn}
%% =========================================================

\subsection{XctNet: Attention-Based Single-View Reconstruction}
\label{sec:cnn:xctnet}

Tan et al.~\cite{xctnet} proposed XctNet, introducing: (1)~a self-attention
module for adaptive feature optimization; (2)~a multi-scale feature fusion
module integrating representations at multiple spatial resolutions; and
(3)~a 3D branch generation module producing volumetric outputs at different
generation fields. XctNet achieved SSIM of 0.8681 and PSNR of 29.28\,dB,
significantly outperforming state-of-the-art methods at publication time.

\subsection{X-CTRSNet: Joint Reconstruction and Segmentation}
\label{sec:cnn:xctrs}

Ge et al.~\cite{xctrsnet} proposed X-CTRSNet, the first framework for
simultaneous 3D reconstruction and segmentation of cervical vertebrae CT
directly from orthogonal 2D AP and lateral X-ray images. It couples SpaDRNet
(reconstruction) and MulSISNet (segmentation) through a Reconstruction-
Segmentation Consistency (RSC) learning strategy:
\begin{equation}
\mathcal{L}_\text{RSC} = \mathcal{L}_\text{recon}
+ \lambda\,\mathcal{L}_\text{seg}
+ \mu\,\mathcal{L}_\text{consistency}
\label{eq:rsc}
\end{equation}
achieving PSNR=24.58\,dB, SSIM=0.749, and Dice coefficient of 80.44\%.

\subsection{AI-Driven CT Reconstruction: Clinical Perspectives}
\label{sec:cnn:clinical}

Pisuchpen et al.~\cite{pisuchpen2026} provide a comprehensive review of AI in
CT image reconstruction for abdominal imaging. Deep Learning-based Reconstruction
(DLR) techniques leverage CNNs for high-fidelity denoising, enabling
diagnostic-quality CT at substantially reduced radiation doses. These clinical
advances establish the benchmark that synthetic CT from X-ray must meet for
acceptance in radiology practice, and validate the architectural principles
(CNNs for volumetric feature learning, perceptual losses, multi-scale
processing) adopted in the synthesis literature.

%% =========================================================
\section{Comparative Analysis Across Method Categories}
\label{sec:comparison}
%% =========================================================

\subsection{Architectural Paradigm Evolution}
\label{sec:comparison:arch}

Figure~\ref{fig:arch_evolution} illustrates the architectural progression from
classical models to modern generative frameworks. The evolution follows a clear
trajectory: from handcrafted geometric models requiring explicit anatomical
knowledge, through discriminative CNN encoders, to generative frameworks
modelling the full distribution of plausible 3D volumes.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.82,every node/.style={font=\scriptsize}]
  \draw[very thick,-{Stealth}] (0,0) -- (0,5.3)
    node[left,rotate=90,font=\small\bfseries,xshift=1.5cm]
    {Accuracy};
  \draw[very thick,-{Stealth}] (0,0) -- (7.3,0)
    node[below,font=\small\bfseries] {Data \& Model Complexity};
  \node[blockO,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (0.9,1.0) {SSM\\ (1990s)};
  \node[blockO,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (2.0,1.7) {Para-\\ metric};
  \node[blockO,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (3.0,2.4) {Contour\\ FFD};
  \node[blockG,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (4.0,3.1) {Early\\ CNN};
  \node[blockR,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (5.0,3.7) {GAN\\ (X2CT)};
  \node[blockR,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (5.8,4.2) {Trans.\\ GAN};
  \node[blockP,text width=1.3cm,minimum height=0.75cm,
    rounded corners=5pt] at (6.6,4.8) {Diff. /\\ NeRF};
  \draw[very thick,dashed,draw=gray!50,-{Stealth}]
    (0.6,0.6) .. controls (3.5,2.8) .. (6.8,5.0);
\end{tikzpicture}
\caption{Qualitative illustration of architectural paradigm evolution in 2D X-ray
to 3D CT reconstruction. Methods are plotted by data/model complexity (x-axis)
and reconstruction quality (y-axis), showing a consistent upward trend.}
\label{fig:arch_evolution}
\end{figure}

\subsection{Input Requirements and View Flexibility}
\label{sec:comparison:input}

Classical methods typically require two or more calibrated X-ray views with
explicit correspondence annotations. Early CNNs and GANs use biplanar inputs.
Recent methods---XctNet, AP2CT-GAN, PTX2CT-GAN, MedNeRF, and X2BR---push the
frontier of single-view reconstruction, which is the most clinically practical
scenario. DX2CT handles both mono- and biplanar inputs within a unified
framework, providing flexibility for different clinical settings.

\subsection{Anatomical Scope}
\label{sec:comparison:anatomy}

Table~\ref{tab:anatomical_scope} summarizes the anatomical coverage across
methods. Classical methods were predominantly validated on skeletal structures
(femur, tibia, pelvis, spine). Modern deep learning methods have substantially
broadened the anatomical scope to include chest, lungs, cervical vertebrae, and
pathological conditions (pneumothorax).

\begin{table}[h]
\caption{Anatomical Scope of Representative Methods}
\label{tab:anatomical_scope}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}p{1.7cm}p{2.2cm}p{2.3cm}@{}}
\toprule
\textbf{Region} & \textbf{Classical} & \textbf{DL / Generative}\\
\midrule
Femur / Knee & Karade, Koh, Zhu, Kim & Kasten, DP-GAN+B\\
Pelvis & Lamecker, Zheng & ---\\
Spine & Humbert, Cresson & X-CTRSNet\\
Chest / Lung & --- & X2CT-GAN, TRCT-GAN, AP2CT-GAN, DVG-Diff\\
C-Vertebra & --- & X-CTRSNet\\
Wrist / Bone & --- & Yoon (Diff.), X2BR\\
Pneumothorax & --- & PTX2CT-GAN\\
General & --- & XctNet, MedNeRF\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-Era Quantitative Comparison}
\label{sec:comparison:quant}

Table~\ref{tab:crossera} provides a cross-era comparison of representative
methods, highlighting the progressive improvement in quantitative metrics.
A significant challenge in comparing methods across the literature is the
heterogeneity of evaluation metrics: classical methods report geometric error
metrics (mean point-to-surface distance, RMS error in mm) while modern methods
report image quality metrics (SSIM, PSNR) on standardized datasets.

\begin{table*}[t]
\caption{Cross-Era Comparison of Key Methods Across All Four Eras}
\label{tab:crossera}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}lllllp{2.5cm}p{3.0cm}@{}}
\toprule
\textbf{Method} & \textbf{Era} & \textbf{Type} & \textbf{Region} &
\textbf{Input} & \textbf{Metric} & \textbf{Key Result}\\
\midrule
Karade et al.~\cite{karade2015} & Classical & Contour+LSD & Femur &
  Biplanar & MP2S (mm) & 1.2\,mm; 49\,sec\\
Kasten et al.~\cite{kasten2020} & Early CNN & CNN & Knee &
  Biplanar DRR & Dice / Chamfer & Dice=0.906; Chamfer=1.778\,mm; 0.5\,sec\\
X-CTRSNet~\cite{xctrsnet} & CNN+GAN & Joint R+S & C-Vertebra &
  Biplanar & PSNR / SSIM / Dice & 24.58\,dB; 0.749; 80.44\%\\
XctNet~\cite{xctnet} & CNN & Attn-CNN & General &
  Single & PSNR / SSIM & 29.28\,dB; 0.8681\\
TRCT-GAN~\cite{trctgan} & GAN & Trans.-GAN & Chest &
  Biplanar & SSIM / PSNR & Superior to X2CT-GAN baseline\\
MedNeRF~\cite{mednerf} & NeRF & Neural Implicit & Chest/Knee &
  Single/few & SSIM / qual. & High-fidelity CT projection synthesis\\
DVG-Diffusion~\cite{dvgdiffusion} & Diffusion & Latent Diff. & Chest &
  Few-view & SSIM / PSNR / FID & SOTA; best fidelity-perceptual balance\\
X2BR~\cite{x2br} & Neural Implicit & ConvNeXt+CPD & Bone &
  Single & IoU / Chamfer-L1 & X2B: IoU=0.952; superior anatomical realism\\
\bottomrule
\end{tabular}
\end{table*}

%% =========================================================
\section{Challenges and Future Directions}
\label{sec:challenges}
%% =========================================================

Figure~\ref{fig:challenges_overview} summarizes the primary open challenges.

\begin{figure}[h]
\centering
\begin{tikzpicture}[every node/.style={font=\scriptsize}]
  \node[blockR,text width=1.8cm,minimum height=0.9cm,rounded corners=5pt]
    (center) at (0,0) {\textbf{Open}\\ \textbf{Challenges}};
  \foreach \angle/\lbl/\col in {
    90/Dataset\\ Scarcity/blue!60!black,
    30/Anatomical\\ Generalization/green!60!black,
    330/Resolution\\ \& Detail/orange!70!black,
    270/Clinical\\ Validation/red!60!black,
    210/Workflow\\ Integration/purple!60!black,
    150/Physics-\\ Informed/teal!60!black}{
    \node[draw=\col,fill=\col!10,rounded corners,text width=1.5cm,
      minimum height=0.75cm,text centered,text=\col]
      (\angle) at (\angle:2.4cm) {\lbl};
    \draw[-{Stealth[length=4pt]},\col!70] (center) -- (\angle);
  }
\end{tikzpicture}
\caption{Six key open challenge areas in 2D X-ray to 3D CT reconstruction,
each representing a significant research gap requiring dedicated investigation.}
\label{fig:challenges_overview}
\end{figure}

\subsection{Dataset Scarcity and Ethical Constraints}
\label{sec:challenges:data}

The fundamental bottleneck for all data-driven methods is the availability of
paired X-ray and CT data from the same patient. This is ethically and
practically challenging---subjecting patients to additional radiation purely for
research is generally not justified. Most studies rely on synthetic X-rays
generated from CT data using digitally reconstructed radiograph (DRR)
algorithms, but the domain gap between simulated and real X-ray appearance
limits generalization. Future work should explore: (1)~federated learning pooling
data across institutions; (2)~physics-based X-ray simulation with realistic noise
and scatter; (3)~self-supervised and semi-supervised learning paradigms; and
(4)~foundation model approaches pre-trained on large heterogeneous medical image
datasets.

\subsection{Anatomical Generalization}
\label{sec:challenges:anatomy}

Most current methods are trained and validated on a single anatomical region and
do not generalize to other body parts. The development of universal anatomical
reconstruction models handling the full range of clinical X-ray examinations
remains an open problem. Promising directions include multi-task learning with
shared encoders and region-specific decoders, hierarchical models that first
segment the region of interest, and foundation models trained on large-scale
multi-anatomy datasets with anatomical conditioning.

\subsection{Resolution and Fine Structural Detail}
\label{sec:challenges:resolution}

Current methods achieve promising results at moderate resolution ($128^3$ or
$256^3$ voxels) but struggle to reproduce fine anatomical detail---trabecular
bone microstructure, small pulmonary nodules, delicate vascular structures.
Cascade architectures generating a low-resolution coarse volume followed by
super-resolution, and patch-based training (as in Yoon et al.~\cite{yoon2024}),
represent promising approaches. Implicit neural representations offer theoretical
resolution independence but face optimization challenges at fine scales.

\subsection{Clinical Validation and Regulatory Pathway}
\label{sec:challenges:clinical}

The gap between research validation and clinical deployment is perhaps the most
critical challenge. Current methods are evaluated on quantitative image quality
metrics (SSIM, PSNR, Dice) that correlate imperfectly with clinical utility.
Future work must develop: (1)~task-specific evaluation protocols measuring
radiologist performance in detecting specific pathologies; (2)~prospective
clinical studies validating diagnostic accuracy; (3)~uncertainty quantification
methods allowing clinicians to identify regions of low model confidence; and
(4)~regulatory frameworks for AI-synthesized medical images, which currently do
not exist in most jurisdictions.

\subsection{Integration with Clinical Workflow}
\label{sec:challenges:workflow}

For any reconstruction method to achieve clinical adoption, it must integrate
with existing radiology workflows: inference times of seconds or minutes,
integration with PACS and DICOM standards, robust performance across scanner
vendors, and interpretable outputs. Lightweight architectures (DP-GAN+B's 44\%
parameter reduction) and efficient diffusion samplers represent steps toward
practical deployment. Edge deployment on portable X-ray machines opens new
possibilities for point-of-care 3D imaging in resource-limited settings.

\subsection{Physics-Informed and Anatomically Constrained Models}
\label{sec:challenges:physics}

Pure data-driven methods can produce anatomically implausible
reconstructions---missing organs, distorted geometry, hallucinated
structures---when extrapolating beyond the training distribution. Physics-informed
neural networks (PINNs) can encode X-ray attenuation physics into the network.
Biomechanical constraints (as in X2BR's template-guided registration) ensure
physically plausible geometry. Integrating these constraints with diffusion
models---whose iterative refinement is amenable to constraint injection at each
denoising step---is a highly promising research direction.

%% =========================================================
\section{Conclusion}
\label{sec:conclusion}
%% =========================================================

This survey has traced the evolution of 2D X-ray to 3D CT reconstruction across
more than three decades, from classical statistical shape models and geometric
registration methods through the deep learning revolution and into the modern
era of generative adversarial networks, diffusion models, and neural radiance
fields. Building on the foundational taxonomy of Maken and Gupta~\cite{maken2023},
we have systematically reviewed eleven papers representing the state of the art
as of 2025--2026, organized into four categories: GAN-based methods, diffusion-based
methods, neural radiance field methods, and complementary CNN architectures.

Key insights from this survey are as follows. First, the field has undergone a
clear paradigm shift from handcrafted geometric models to end-to-end learned
generative frameworks, broadening anatomical scope, reducing input requirements
from many calibrated views to a single X-ray, and dramatically improving
reconstruction speed and quality. Second, GAN-based methods have established
biplanar and single-view chest reconstruction as mature capabilities, while
disease-specific adaptations (PTX2CT-GAN) demonstrate the value of targeted
architectural design. Third, diffusion models offer stable training, better
distributional coverage, and built-in uncertainty quantification---DX2CT and
DVG-Diffusion represent current state of the art. Fourth, neural radiance fields
(MedNeRF, X2BR) introduce continuous implicit volumetric representations enabling
extremely sparse-view reconstruction with significant radiation dose reduction
potential. Fifth, hybrid methods combining neural implicit learning with
anatomical priors (X2BR) provide the most clinically grounded reconstructions.

Despite these advances, critical challenges remain: dataset scarcity, anatomical
generalization across body regions, clinical validation through task-specific
evaluation and prospective studies, and regulatory frameworks for AI-synthesized
medical images. Addressing these challenges will require collaboration across the
deep learning research community, medical imaging, clinical radiology, regulatory
agencies, and healthcare systems worldwide.

The vision motivating this field---providing 3D anatomical information at X-ray
radiation doses and costs---is increasingly within reach. As methods reviewed
in this survey continue to mature and clinical validation studies accumulate,
X-ray to CT reconstruction has the potential to transform diagnostic imaging
practice, bringing CT-level diagnostic power to settings that currently lack
access while reducing radiation exposure for all patients.

%% =========================================================

%% =========================================================


%% =========================================================
%% REFERENCES
%% =========================================================
\begin{thebibliography}{99}

\bibitem{maken2023}
Maken P, Gupta A (2023) 2D-to-3D: A review for computational 3D image
reconstruction from X-ray images. Arch Comput Methods Eng 30:85--114.
\url{https://doi.org/10.1007/s11831-022-09790-z}

\bibitem{goodfellow2014}
Goodfellow I, Pouget-Abadie J, Mirza M et al (2014) Generative adversarial
nets. In: Advances in Neural Information Processing Systems (NeurIPS) 27

\bibitem{ho2020ddpm}
Ho J, Jain A, Abbeel P (2020) Denoising diffusion probabilistic models.
NeurIPS 33:6840--6851

\bibitem{nerf2020}
Mildenhall B, Srinivasan PP, Tancik M et al (2020) NeRF: Representing scenes
as neural radiance fields for view synthesis. ECCV 2020. Lect Notes Comput Sci
12346:405--421

\bibitem{cootes1992}
Cootes TF, Taylor CJ, Cooper DH, Graham J (1992) Training models of shape from
sets of examples. Proc BMVC 1992:9--18

\bibitem{kasten2020}
Kasten Y, Doktofsky D, Kovler I (2020) End-to-end convolutional neural network
for 3D reconstruction of knee bones from bi-planar X-ray images. In: Mach Learn
Med Image Reconstr (MLMIR 2020). Springer, pp 123--133

\bibitem{dixit2019}
Dixit S, Pai VG, Rodrigues VC et al (2019) 3D reconstruction of 2D X-ray
images. In: Proc 4th Int Conf CSITSS 2019, pp 1--5

\bibitem{kim2019}
Kim H, Lee K, Lee D, Baek N (2019) 3D reconstruction of leg bones from X-ray
images using CNN-based feature analysis. In: Proc ICTC 2019, pp 669--672

\bibitem{lamecker2006}
Lamecker H, Wenckebach T, Hege H-C (2006) Atlas-based 3D-shape reconstruction
from X-ray images. In: Proc ICPR 2006. IEEE, pp 371--374

\bibitem{zheng2010}
Zheng G (2010) Statistical shape model-based reconstruction of a scaled,
patient-specific surface model of the pelvis from a single standard AP X-ray
radiograph. Med Phys 37:1424--1439

\bibitem{zhu2011}
Zhu Z, Li G (2011) Construction of 3D human distal femoral surface models
using a 3D statistical deformable model. J Biomech 44:2362--2368

\bibitem{humbert2009}
Humbert L, De Guise JA, Aubert B, Godbout B, Skalli W (2009) 3D reconstruction
of the spine from biplanar X-rays using parametric models. Med Eng Phys
31:681--687

\bibitem{chaibi2012}
Chaibi Y, Cresson T, Benjamin A et al (2012) Fast 3D reconstruction of the
lower limb using a parametric model from biplanar X-rays. Comput Methods
Biomech Biomed Eng 15:457--466

\bibitem{koh2011}
Koh K, Kim YH, Kim K, Park WM (2011) Reconstruction of patient-specific
femurs using X-ray and sparse CT images. Comput Biol Med 41:421--426

\bibitem{gunay2007}
Gunay M, Shim MB, Shimada K (2007) Cost- and time-effective three-dimensional
bone-shape reconstruction from X-ray images. Int J Med Robot.
\url{https://doi.org/10.1002/rcs.162}

\bibitem{karade2015}
Karade V, Ravi B (2015) 3D femur model reconstruction from biplane X-ray images:
a novel method based on Laplacian surface deformation. Int J Comput Assist
Radiol Surg 10:473--485

\bibitem{x2ctgan}
Ying X, Guo H, Ma K et al (2019) X2CT-GAN: Reconstructing CT from biplanar
X-rays with generative adversarial networks. In: Proc CVPR 2019, pp 10619--10628

\bibitem{cvaegan}
Jiang L, Zhang M, Wei R et al (2021) Reconstruction of 3D CT from a single
X-ray projection view using CVAE-GAN. In: Proc IEEE ICMIPE 2021.
\url{https://doi.org/10.1109/ICMIPE53131.2021.9698875}

\bibitem{xctrsnet}
Ge R, He Y, Xia C et al (2022) X-CTRSNet: 3D cervical vertebra CT
reconstruction and segmentation directly from 2D X-ray images. Knowl Based
Syst 236:107680. \url{https://doi.org/10.1016/j.knosys.2021.107680}

\bibitem{xctnet}
Tan Z, Li J, Tao H, Li S, Hu Y (2022) XctNet: Reconstruction network of
volumetric images from a single X-ray image. Comput Med Imaging Graph
98:102067. \url{https://doi.org/10.1016/j.compmedimag.2022.102067}

\bibitem{trctgan}
Wang Y, Sun Z-L, Zeng Z, Lam K-M (2023) TRCT-GAN: CT reconstruction from
biplane X-rays using transformer and generative adversarial networks. Digit
Signal Process 140:104123. \url{https://doi.org/10.1016/j.dsp.2023.104123}

\bibitem{dpganb}
Xing X, Li X, Wei C et al (2024) DP-GAN+B: A lightweight generative adversarial
network based on depthwise separable convolutions for generating CT volumes.
Comput Biol Med 174:108393.
\url{https://doi.org/10.1016/j.compbiomed.2024.108393}

\bibitem{yoon2024}
Yoon S, Pratap JS, Liu W-C et al (2024) High-resolution 3D CT synthesis from
bidirectional X-ray images using 3D diffusion model. In: Proc IEEE ISBI 2024.
\url{https://doi.org/10.1109/ISBI56570.2024.10635648}

\bibitem{dx2ct}
Jeong YS, Yoo HB, Chun IY (2025) DX2CT: Diffusion model for 3D CT
reconstruction from bi or mono-planar 2D X-ray(s). In: Proc IEEE ICASSP 2025.
\url{https://doi.org/10.1109/ICASSP49660.2025.10888986}

\bibitem{ap2ctgan}
Tang T, Cai W, Yang B et al (2025) A novel medical image reconstruction
framework for 3D chest CT based on a single 2D anteroposterior X-ray image.
In: Proc IEEE SMC 2025.
\url{https://doi.org/10.1109/SMC58881.2025.11342980}

\bibitem{ptx2ctgan}
Wang Q, Tang T, Huang J et al (2025) PTX2CT-GAN: Reconstruction of chest CT
from single X-ray radiography for pneumothorax. In: Proc IEEE BIBM 2025.
\url{https://doi.org/10.1109/BIBM66473.2025.11356273}

\bibitem{dvgdiffusion}
Xie X, Liu J, Fan H et al (2026) DVG-Diffusion: Dual-view-guided diffusion
model for CT reconstruction from X-rays. IEEE Trans Image Process 35:1158--1172

\bibitem{mednerf}
Corona-Figueroa A, Frawley J, Bond-Taylor S et al (2022) MedNeRF: Medical
neural radiance fields for reconstructing 3D-aware CT-projections from a single
X-ray. In: Proc IEEE EMBC 2022.
\url{https://doi.org/10.1109/EMBC48229.2022.9871757}

\bibitem{x2br}
Guven G, Ugurdag F, Ates HF (2026) X2BR: High-fidelity 3D bone reconstruction
from a planar X-ray image with hybrid neural implicit methods. Med Biol Eng
Comput. \url{https://doi.org/10.1007/s11517-026-03549-5}

\bibitem{pisuchpen2026}
Pisuchpen N, Srinivas Rao S, Noda Y et al (2026) Artificial intelligence (AI)
and CT in abdominal imaging: image reconstruction and beyond. Abdom Radiol
51:424--445. \url{https://doi.org/10.1007/s00261-025-05031-6}

\bibitem{gupta2020}
Gupta A (2020) Challenges for computer aided diagnostics using X-ray and
tomographic reconstruction images in craniofacial applications. Int J Comput
Vis Robot 10:11. \url{https://doi.org/10.1504/IJCVR.2020.10029170}

\end{thebibliography}

\end{document}